Automatically proving equivalence by type-safe reflection

### 2016-01-19 paper-JFP-DTP/content/1-introduction.tex
new fix

\subsection{Propositional and Judgemental Equalities in Intentional Type Theories}
In Mathematics, equality is a proposition, which can hold or otherwise, or be disproved.
Since in type theory, propositions are seen as types~\cite{How80}, the
proposition that two elements $x$ and $y$ are equal corresponds to a type.
Thus, if $x$ and $y$ are of type $a$, then the type
$Id_a(x, y)$ represents the proposition ``$x$ is equal to $y$". If this type is
inhabited, then $x$ is said to be provably equal to $y$.  Thus, $Id$ is a type
family (parametrised by the type $a$) indexed over two elements of $a$, giving
$Id (a:Type)\ :\ a \rightarrow a \rightarrow Type$. For
convenience, we write $Id_a\ x\ y$ as $x\ =_a\ y$.  This equality is the
equality which can be manipulated in the language.

There is another more
primitive notion of equality in Intentional Type Theories, called judgemental
equality, or definitional equality. This second equality means ``equal by
definition'' The judgemental equality cannot be negated or
assumed; we cannot talk about this primitive equality inside the theory.
Whether or not two expressions are equal by definition is a matter of
evaluating the definitions. For example, if $f\ :\ \mathbb{N}\ \rightarrow\
\mathbb{N}$ is defined by $f\ x\ \equiv\ x\ +\ 2$, then $f\ 5$ is
definitionally equal to $7$. Definitional equality entails unfolding
of functions and reductions, until no more reduction can be
performed. We denote definitional equality by $\equiv$.

@@

The only way for $(Id_a\ x\ y)$ to be inhabited is therefore that $x$ should be
(by definition!) $y$. In this case, the constructor $Refl$ helps to create a
proof of this equality : $Refl\ x$ is precisely the proof which says that
$x=_ax$.

@@

It should be pointed that the proof component of this function is established without any effort, as we always give the ``Refl" for the proof. Indeed, for each of the eight patterns, a simple computation reduces the given and expected values to the same value, and ``Refl" can therefore be used. We also note that
there is no need to produce any lemma of correctness about $addBit$ afterwards.
The correct by construction style in which it is written already gives the
property that we want: the two bits produced effectively represent the
addition of the three bits given in input.

We finally define the function adding two binary numbers and a carry flag.
This addition works for two binary numbers
expressed with the same number of bits, and it produces a result with one more
bit. This result represents the value $c + x + y$, where $c$ 
representis the value of the input carry bit, and $x$ and $y$ represent
the values of the two binary numbers given in input. Thus, we'd like to
be able to write:

@@

The principal novelty is in the approach that we follow, using a
new kind of \emph{type-safe reflection}.  Working by reflection for implementing
tactics has been done several times, for the implementation of a
ring solver for the Coq proof assistant, but without
providing any guarantees, contrary to our type-safe reflection. We will
compare our approach with other implementations in Section~\ref{sect:discuss}.

@@

        \item We develop a hierarchy of tactics where each tactic reuses the
        rewriting machinery from the structure immediately above in the hierarchy.
        For example,
        simplifying neutral elements is only implemented at the monoid level,
        and each level above monoid reuses it. Re-usability can be quite tricky
        to obtain when we want to reuse a prover from a structure which is less
        expressive. For example, reusing the monoid prover for building the
        group prover is not trivial, since we lose the possibility to express
        negations ($-x$) and subtractions ($x-y$). Some specific encodings
        will be presented in Section~\ref{sect:reuse} for solving this problem.

>>>>

\subsection{Propositional and Judgemental Equalities in Intentional Type Theories}
In Mathematics, equality is a proposition, e.g. we can disprove an equality or 
assume an equality as a hypothesis.
Since in type theory, propositions are seen as types~\cite{How80}, the
proposition that two elements $x$ and $y$ are equal corresponds to a type.
Thus, if $x$ and $y$ are of type $a$, then the type
$Id_a(x, y)$ represents the proposition ``$x$ is equal to $y$". If this type is
inhabited, then $x$ is said to be provably equal to $y$.  Thus, $Id$ is a type
family (parametrised by the type $a$) indexed over two elements of $a$, giving
$Id (a:Type)\ :\ a \rightarrow a \rightarrow Type$. For
convenience, we write $(Id_a\ x\ y)$ as $(x\ =_a\ y)$. This equality is the
equality which can be manipulated in the language.

There is another, more
primitive, notion of equality in Intentional Type Theories, called judgemental
equality, or definitional equality. This second equality means ``equal by
definition'' The judgemental equality cannot be negated or
assumed; we cannot talk about this primitive equality inside the theory.
Whether or not two expressions are equal by definition is a matter of
evaluating the definitions. For example, if $f\ :\ \mathbb{N}\ \rightarrow\
\mathbb{N}$ is defined by $f\ x\ \equiv\ x\ +\ 2$, then $f\ 5$ is
definitionally equal to $7$. Definitional equality entails unfolding
of functions and reductions, until no more reduction can be
performed. We denote the definitional equality by $\equiv$.

@@

The only way for $(Id_a\ x\ y)$ to be inhabited is therefore that $x$ should be
(by definition!) $y$. In this case, the constructor $Refl$ helps to create a
proof of this equality : $(Refl\ x)$ is precisely the proof which says that
$x=_ax$.

@@

It should be pointed that the proof component of this function is established without any effort, as we always give ``Refl" for the proof. Indeed, for each of the eight patterns, a simple computation reduces the given and expected values to the same value, and ``Refl" can therefore be used. We also note that
there is no need to produce any lemma of correctness about $addBit$ afterwards.
The correct by construction style in which it is written already gives the
property that we want: the two bits produced effectively represent the
addition of the three bits given in input.

We finally define the function adding two binary numbers and a carry flag.
This addition works for two binary numbers
expressed with the same number of bits, and it produces a result with one more
bit. This result represents the value $c + x + y$, where $c$ 
represents the value of the input carry bit, and $x$ and $y$ represent
the values of the two binary numbers given in input. Thus, we'd like to
be able to write:

@@

The principal novelty is in the approach that we follow, using a
new kind of \emph{type-safe reflection}.  Working by reflection for implementing
tactics has been done several times, including for the implementation of a
ring solver for the Coq proof assistant, but without
providing any guarantees, contrary to our type-safe reflection. We will
compare our approach with other implementations in Section~\ref{sect:discuss}.

@@

        \item We develop a hierarchy of tactics where each tactic reuses the
        rewriting machinery from the structure immediately below in the hierarchy.
        For example,
        simplifying neutral elements is only implemented at the monoid level,
        and each level above monoid reuses it. Re-usability can be quite tricky
        to obtain when we want to reuse a prover from a structure which is less
        expressive. For example, reusing the monoid prover for building the
        group prover is not trivial, since we lose the possibility to express
        negations ($-x$) and subtractions ($x-y$). Some specific encodings
        will be presented in Section~\ref{sect:reuse} for solving this problem.

### 2016-01-19 paper-JFP-DTP/content/2-basicIdeas.tex
new fix

If we continue with this first definition of $Expr$, the normalisation function will certainly take an Expr and produce another Expr, and we will need to prove the following correctness lemma afterwards : \\
$\forall\ e:Expr\ \Gamma,\ reify\ (reduce\ e)\ =\ reify\ e$ \\
where $reify$ is a function computing the interpretation of an Expr in a context $\Gamma$, that is to say, the natural number that this Expr is encoding.
This proof can be quite tricky to build because it relies on the complete behaviour of the reduction and on the way the interpretation is computed.
For this little example, the reduction procedure won't be too heavy, but in the next sections, with more sophisticated algebraic structures, we will have more properties to deal with and it will certainly become more problematic to unfold the definition of a gigantic reduction procedure. We want to avoid pain as much as possible, and for this reason, we don't want to have to write big proofs, and even to avoid the writing of proofs as much as possible. \\
\\
To avoid this source of complexity, we add an index to the type $Expr$, and this index is precisely the concrete number that the Expr is encoding. This is the first brick to our type-safe reflection mechanism. Thus, it won't be necessary to define the $reify$ function, as we will know directly the concrete element reflected by a term of type $Expr\ \Gamma\ x$ just by looking at its index $x$. We also get the guarantee that the reflection of an expression $e$ is indeed a faithful representation of $e$ \\

>>>>

If we continue with this first definition of $Expr$, the normalisation function will certainly take an Expr and produce another Expr, and we will need to prove the following correctness lemma afterwards : \\
$\forall\ e:Expr\ \Gamma,\ reify\ (reduce\ e)\ =\ reify\ e$ \\
where $reify$ is a function computing the interpretation of an Expr in a context $\Gamma$, that is to say, the natural number that this Expr is encoding.
This proof can be quite tricky to build because it relies on the complete behaviour of the reduction and on the way the interpretation is computed.
For this little example, the reduction procedure won't be too heavy, but in the next sections, with more sophisticated algebraic structures, we will have more properties to deal with and it will certainly become more problematic to unfold the definition of a gigantic reduction procedure. We want to avoid pain as much as possible, and for this reason, we want to avoid the writing of big proofs, and even to avoid the writing of proofs as much as possible. \\
\\
To avoid this source of complexity, we add an index to the type $Expr$, and this index is precisely the concrete number that the Expr is encoding. This is the first brick to our type-safe reflection mechanism. Thus, it won't be necessary to define the $reify$ function, as we will know directly the concrete element reflected by a term of type $Expr\ \Gamma\ x$ just by looking at its index $x$. We also get the guarantee that the reflection of an expression $e$ is indeed a faithful representation of $e$ \\
